Covering Basis: Comprehensive Model Agnostic Probes of the Binary Black Hole Population

### 2022-10-09 src/tex/appendicies.tex
Move some stuff around in prep for new results -- dont ever mention mspline

\appendix
\section{Penalized Splines and Smoothing}\label{sec:psplines}

Spline functions have been shown to be sensitive to the chosen number of knots, and their locations or spacing \citep{deBoor78}. 
Adding more knots increases the a priori variance in the spline function, while the space between knots can limit the 
resolution of features in the data the spline is capable of resolving. To ensure your spline based model is flexible enough 
one would want to add as many knots as densely as possible, but this comes with unwanted side effect of larger variance as mentioned above. 
To overcome this we resort to the frequentest penalized spline (P-Spline) in which one applies a penalty or regularization term 
to the likelihood based on the difference of adjacent knot coefficients \citep{eilers2021practical}. The linear combination of spline basis components 
or the resulting spline function is flat when the basis coefficients are equal (see Figure \ref{fig:spline_basis}). By penalizing the likelihood as 
the differences between adjacent knot coefficients get larger, one gets a smoothing of the spline function inferences \citep{eilers2021practical}. 
Since we are using hierarchical Bayesian inference as our statistical framework, we formulate this penalized likelihood with its Bayesian analog. 
The difference penalty is equivalent to applying a stochastic difference prior among the knots, as described in \citet{BayesianPSplines} and \citet{Jullion2007RobustSO}. 
This is also sometimes referred to as a Gaussian random walk prior, since each coefficient is normally distributed about its neighboring coefficients. 
For a spline basis with $n$ degree's of freedom, and a difference penalty of order of $r$ (see \citet{eilers2021practical}), the smoothing prior is defined as:

@@

\noindent Above $\bm{D}_r$ is the order-$r$ difference matrix, of shape $(n-r \times n)$ and $\Gamma(a,b)$ the Gamma distribution with shape parameters, $a$ and $b$. 
This smoothing prior removes the strong dependence on number and location of knots that arises with using splines. The $\tau_\lambda$ controls the ``strength'' of 
smoothing, or the inverse variance of the Gaussian priors on knot differences. By adding in a very large number 
of knots such that your domain is densely populated with basis coefficients, this allows the freedom for the model to find the smoothing 
scale that the data prefers. One downside of this prior specification as noted in \citet{BayesianPSplines} and \citet{Jullion2007RobustSO}, is a slight dependence on choice 
of the b hyperparameter for Gamma distribution prior on $\tau_\lambda$. We reduce this affect by following the method introduced in \citet{Jullion2007RobustSO}
to alleviate this. This is done by modeling $\tau_\lambda$ with a mixture model on gamma distributions with varying b hyperparameters. 
We now replace the prior on $\tau_\lambda$ with:

@@

\noindent Where above $\mathcal{D}(m)$ is a Dirichlet distribution with $m$ degrees of freedom and ones as the concentration parameters, 
and the $b_i$'s are $m$ different b shape parameters for the Gamma distribution.

@@

Since the prior is imparting a natural attraction of the coefficients closer to each other in order to smooth the spline function, so one 
must ensure that the spline function is in fact flat given all equal coefficients. Since one needs to define n+k knots to construct an order-k basis with 
n degrees of freedom, sometimes studies place knots on top of each other at hard parameter boundaries \cite{deBoor78,monotone_regression_splines}. 
This may seem motivated, especially when is no reality in the parameter beyond the boundary, but this violates the above condition. 
We follow the distinction from \citet{eilers2021practical} that such a smoothing prior is only valid with ``proper'' spline bases. 
A proper basis is where all n+k knots are evenly spaced, see Figure \ref{fig:spline_basis}, as opposed to stacking them at the bounds.

>>>>

\appendix
\section{Basis Splines} \label{sec:basis_splines}

A common non-parametric method used in many statistical applications is basis splines. A spline function of order $k$, 
is a piece-wise polynomial of order $k$ polynomials stitched together from defined ``knot'' locations across the domain. 
They provide a useful and cheap way to interpolate generically smooth functions from a finite sampling of ``knot'' heights. 
Basis splines of order $k$ are a set of order $k$ polynomials that form a complete basis for any spline function of order $k$. 
Therefore, given an array of knot locations, $\mathbf{t}$ or knot vector, there exists a single unique linear combination of basis splines 
for every possible spline function interpolated from knots, $\mathbf{t}$. The basis components for a given knot vector can be constructed  
using the Cox-de Boor recursion formula \citep{deBoor78,monotone_regression_splines}. Given knots $t_0$, $t_1$,...,$t_{i+k}$ 
the base case and recursion relation that generates this particular basis are defined as:

\begin{equation}
    B_{i,1}(x | \mathbf{t}) = 
    \begin{cases}
        1, & \text{if } t_i \leq x < t_{i+1} \\
        0, & \text{otherwise}
    \end{cases}
\end{equation}

\begin{equation}
    B_{i,k+1}(x | \mathbf{t}) = \omega_{i,k}(x | \mathbf{t})B_{i,k}(x | \mathbf{t}) + \big[1-\omega_{i+1,k}(x | \mathbf{t})\big] B_{i+1,k}(x | \mathbf{t})
\end{equation}

\begin{equation}
    \omega_{i,k}(x | \mathbf{t}) =
    \begin{cases}
        \frac{x-t_i}{t_{i+k}-t_i}, & t_{i+k} \neq t_i \\
        0, & \text{otherwise}
    \end{cases}
\end{equation}

\noindent This is known as the ``B-Spline'' basis after it's inventor de Boor \citep{deBoor78}. The power of basis splines
comes from the fact that one only has to do the somewhat-expensive interpolation once for each set of points at which the spline is evaluated. 
This provides a considerable computational speedup as each evaluation of the spline function becomes a simpler operation: a dot product of a 
matrix and a vector. This straightforward operation is also ideal for optimizations from the use of GPU accelerators, 
enabling our Markov chain Monte Carlo (MCMC) based analyses, often with hundreds of parameters, to converge in an hour or less. 
Basis splines can easily be generalized to their two-dimensional analog, producing tensor product basis splines that, 
with this computational advantage, allow for high fidelity modeling of two-dimensional spline functions.

\begin{figure}[ht!]
    \begin{centering}
        \includegraphics[width=\linewidth]{figures/spline_basis_plot.pdf}
        \caption{Plot showing a ``proper'' (see appendix \ref{sec:psplines}) normalized B-Spline basis of order 3 (cubic) with 20 degrees of freedom and equal weights for each component. 
        In black, we show the resulting spline function given equal weights and denote the location of the knots with gray x's.}
        \label{fig:spline_basis}
    \end{centering}
    \script{spline_basis_plot.py}
\end{figure}

Another important feature of basis splines is that under appropriate prior conditions, one can alleviate sensitivities to arbitrarily 
chosen prior specifications that splines commonly struggle with. Previous studies using splines had to perform multiple analyses, varying the 
number of spline knots, then either marginalized over the models or used model comparisons to motivate the best choice \citep{Edelman_2022ApJ}. 
We can avoid this step with the use of penalized splines (or P-Splines) \citep{eilers2021practical,BayesianPSplines,Jullion2007RobustSO}, 
where one adds a smoothing prior comprised of Gaussian distributions on the differences between neighboring basis spline coefficients. 
This prior allows one to populate the domain more densely with knots without the worry of extra variance in the inferred spline functions. When also fitting the scale 
of the smoothing prior (i.e. the width of the Gaussian distributions on the differences), this allows the data to inform the scale of smoothing required. 
We discuss the details of our smoothing prior implementation in more detail in the next section, Appendix \ref{sec:psplines}, 
along with our specific prior and basis choices for each model following that in Appendix \ref{sec:modelpriors}.

\section{Penalized Splines and Smoothing Priors}\label{sec:psplines}

Spline functions have been shown to be sensitive to the chosen number of knots, and their locations or spacing \citep{deBoor78}. 
Adding more knots increases the a priori variance in the spline function, while the space between knots can limit the 
resolution of features in the data the spline is capable of resolving. To ensure your spline based model is flexible enough 
one would want to add as many knots as densely as possible, but this comes with unwanted side effect of larger variance imposed by your model. 
This can be fixed with the use of penalized splines (P-Spline) in which one applies a prior or regularization term 
to the likelihood based on the difference of adjacent knot coefficients \citep{eilers2021practical}. The linear combination of spline basis components 
or the resulting spline function is flat when the basis coefficients are equal (see Figure \ref{fig:spline_basis}). By penalizing the likelihood as 
the differences between adjacent knot coefficients get larger, one gets a smoothing effect on the spline function \citep{eilers2021practical}. 
With hierarchical Bayesian inference as our statistical framework, we formulate the penalized likelihood of \citet{eilers2021practical}'s P-Splines with 
their Bayesian analog \citep{BayesianPSplines}. The Bayesian P-Spline prior places Gaussian distributions over the $r$-th order differences of the 
coefficients \citep{BayesianPSplines,Jullion2007RobustSO}. This is also sometimes referred to as a Gaussian random walk prior. 
For a spline basis with $n$ degree's of freedom, and a difference penalty of order of $r$ (see \citet{eilers2021practical}), 
the smoothing prior is defined as:

@@

\noindent Above $\bm{D}_r$ is the order-$r$ difference matrix, of shape $(n-r \times n)$ and $\Gamma(a,b)$ the Gamma distribution with shape parameters, $a$ and $b$. 
This smoothing prior removes the strong dependence on number and location of knots that arises with using splines. The $\tau_\lambda$ controls the ``strength'' of 
smoothing, or the inverse variance of the Gaussian priors on knot differences. When there are a very large number of knots, 
such that your domain is densely populated with basis coefficients, this allows the freedom for the model to find the smoothing 
scale that the data prefers. One downside of this prior specification as noted in \citet{BayesianPSplines} and \citet{Jullion2007RobustSO}, is the dependence on 
one's choice of value for the $b$ hyperparameter in the Gamma distribution for $\tau_\lambda$. The models introduced in \citet{Jullion2007RobustSO} show how 
to alleviate this. W model $\tau_\lambda$ with a mixture model composed of multiple Gamma distributions each with varying values for $b$. 
With this we now have the full prior on $\tau_\lambda$ as:

@@

\noindent Above $\mathcal{D}(m)$, denotes a Dirichlet distribution with $m$ degrees of freedom and ones as the concentration parameters, 
and the $b_i$'s are $m$ different b shape parameters for the Gamma distribution.

@@

This prior is imparting a natural attraction of the coefficients closer to each other in order to smooth the spline function, so one 
must ensure that the spline function is in fact flat given all equal coefficients. There needs to be $n+k$ knots to construct an order-k 
basis with n degrees of freedom. Some studies place knots on top of each other at hard parameter boundaries \citep{deBoor78,monotone_regression_splines}, 
which may seem motivated, but this violates the above condition necessary for the P-Spline prior. We follow the distinction in \citet{eilers2021practical} 
that such a smoothing prior is only valid with ``proper'' spline bases. A proper basis is where all $n+k$ knots are evenly and equally spaced, 
see Figure \ref{fig:spline_basis}, as opposed to stacking them at the bounds.

### 2022-10-09 src/tex/intro.tex
Move some stuff around in prep for new results -- dont ever mention mspline

Studies using flexible methods to analyze the binary black hole (BBH) mass distribution with GWTC-2 found signs of structure beyond the first reported feature found at 
$m_1\sim35\,\msun$ \citep{Talbot_2018,o3a_pop,Tiwari_2021_b,Edelman_2022ApJ}. The evidence for this structure, 
such as the peak at $m_1\sim10\,\msun$, only grew after analyzing the latest catalog, GWTC-3, with the same models \citep{o3b_astro_dist,Tiwari_2022ApJ}, illustrating  
the usefulness of data-driven methods even with the relatively small current catalog size. Canonically, the approach to constructing population models has been 
to use simple parametric models that aim to describe the data in the simplest way, build in astrophysically motivated priors, then sequentially add in 
complexities only as the data demands. This simple approach is powerful in data-scarce, prior dominated environments; however, as we go forward into 
the data-rich catalog era, flexible methods such as non-parametric modeling will be necessary to discover new features in the data that can quickly 
become unwieldy to incorporate into the parametric approach. As opposed to parametric models, flexible and non-parametric models are data-driven and 
contribute little bias to functional form. Non-parametric or semi-parametric methods are particularly useful to search for unexpected features in the 
data, such as meaningful features that parametric models may fail to capture. Development and robust testing of such flexible methods in the short 
term will only provide knock on effects when the time comes to analyze the next catalog of CBC detections.

>>>>

Studies using flexible methods to analyze the binary black hole (BBH) mass distribution with GWTC-2 found signs of structure beyond the first reported feature found at 
$m_1\sim35\,\msun$ \citep{Talbot_2018,o3a_pop,Tiwari_2021_b,Edelman_2022ApJ}. The evidence for this structure, 
such as the peak at $m_1\sim10\,\msun$, only grew after analyzing the latest catalog, GWTC-3, with the same models \citep{o3b_astro_dist,Tiwari_2022ApJ}, illustrating  
the usefulness of data-driven methods even with the relatively small current catalog size. Canonically, the approach to constructing population models has been 
to use simple parametric models that aim to describe the data in the simplest way, build in astrophysically motivated priors, then sequentially add in 
complexities only as the data demands. This simple approach is powerful in data-scarce, prior dominated environments; however, as we go forward into 
the data-rich catalog era, flexible methods such as non-parametric modeling will be necessary to discover new features in the data that can quickly 
become unwieldy to incorporate into the parametric approach. As opposed to parametric models, flexible and non-parametric models are data-driven and 
contribute little bias to functional form. Non-parametric or semi-parametric methods are particularly useful to search for unexpected features in the 
data, providing meaningful insight of these features that parametric models may fail to capture. Development and robust testing of such flexible methods in the short 
term will only provide knock on effects when the time comes to analyze the next catalog of CBC detections.

### 2022-10-09 src/tex/methods.tex
Move some stuff around in prep for new results -- dont ever mention mspline

\section{Constructing A Basis} \label{sec:basis_splines}

A common non-parametric method used in many statistical applications is basis splines. A spline function of order $k$, 
is a piece-wise polynomial of order $k$ polynomials stitched together from defined ``knot'' locations across the domain. 
They provide a useful and cheap way to interpolate generically smooth functions from a finite sampling of ``knot'' heights. 
Basis splines of order $k$ are a set of order $k$ polynomials that form a complete basis for any spline function of order $k$. 
Therefore, given an array of knot locations, $\mathbf{t}$ or knot vector, there exists a single unique linear combination of basis splines 
for every possible spline function interpolated from knots, $\mathbf{t}$. The basis components for a given knot vector are constructed  
using the Cox-de Boor recursion formula \citep{deBoor78}. Given knots $t_0$, $t_1$,...,$t_{i+k}$ we start with the base case as:

\begin{figure}[ht!]
    \begin{centering}
        \includegraphics[width=\linewidth]{figures/spline_basis_plot.pdf}
        \caption{Plot showing a ``proper'' (see appendix \ref{sec:psplines}) M-Spline basis of order 3 (cubic) with 20 degrees of freedom and equal weights for each component. 
        In black, we show the resulting spline function with given equal weights and denote the location of the knots with gray x's.}
        \label{fig:spline_basis}
    \end{centering}
    \script{spline_basis_plot.py}
\end{figure}

\begin{equation}
    B_{i,1}(x | \mathbf{t}) = 
    \begin{cases}
        1, & \text{if } t_i \leq x < t_{i+1} \\
        0, & \text{otherwise}
    \end{cases}
\end{equation}

\noindent combined with the recursion relation:

\begin{multline*}
    B_{i,k+1}(x | \mathbf{t}) = \omega_{i,k}(x | \mathbf{t})B_{i,k}(x | \mathbf{t})\\
                                + \big[1-\omega_{i+1,k}(x | \mathbf{t})\big] B_{i+1,k}(x | \mathbf{t})
\end{multline*}

\noindent where above $\omega_{i,k}$ is defined as:

\begin{equation}
\omega_{i,k}(x | \mathbf{t}) =
\begin{cases}
    \frac{x-t_i}{t_{i+k}-t_i}, & t_{i+k} \neq t_i \\
    0, & \text{otherwise}
\end{cases}
\end{equation}

\noindent This is known as the ``B-Spline'' basis after it's inventor de Boor \citep{deBoor78}. The power of basis splines
comes from the fact that one only has to do the somewhat-expensive interpolation once for each set of points at which the spline is evaluated. 
This provides a considerable computational speedup as each evaluation of the spline function becomes a simpler operation: a dot product of a 
matrix and a vector. This straightforward operation is also ideal for optimizations from the use of GPU accelerators, 
enabling expensive Markov chain Monte Carlo (MCMC) based analyses with hundreds of parameters \bruce{citation?}. 
Basis splines can easily be generalized to their two-dimensional analog, producing tensor product basis splines that, 
with this computational advantage, allow for high fidelity modeling of two-dimensional spline functions.

A similar basis, the M-Spline basis \citep{monotone_regression_splines}, has desirable properties when looking to model a probability density function, 
which is the goal of population inference. Each basis component is scaled such that it is normalized over the domain, $\int M_i(x)dx = 1$. This ensures 
that given a vector of basis coefficients, if they sum to unity, the resulting spline function is normalized across the interpolation domain. 
The M-Spline basis differs from B-Splines by the scaling factor that normalizes each component:

\begin{equation}\label{eq:MB_SplineRelation}
M_{i,k} = \frac{k}{t_{i+k} - t_i} B_{i,k}
\end{equation}

\noindent We use this definition of the M-Spline basis for all the basis spline models used in this manuscript. Figure \ref{fig:spline_basis} 
shows an example of a cubic (3rd order) M-Spline basis with 20 degrees of freedom and knots linearly spaced from 0 to 1.
Another important feature of basis splines is that under appropriate prior conditions, one can alleviate sensitivities to arbitrarily 
chosen prior specifications that splines commonly struggle with. We look to the use of penalized splines (or P-Splines) \citep{eilers2021practical,BayesianPSplines,Jullion2007RobustSO}, 
where one adds a smoothing prior based on the difference of basis spline coefficients. This allows one to populate the domain with more dense knots 
without the worry of extra variance in the inferred spline functions. We discuss the details of our smoothing prior implementation 
in more detail in Appendix \ref{sec:psplines}, along with our specific prior and basis choices for each model in Appendix \ref{sec:modelpriors}.

>>>>

\section{Building the Model} \label{sec:methods}

We parameterize the binaries' masses with the primary (defined as the more massive component) mass 
and the mass ratio -- defined as $q=\frac{m_2}{m_1}$, bounded in the range from 0 to 1. Furthermore, we model 4 of the 6 total 
spin degrees of freedom of a binary merger with each component spin magnitude $a_1$ and $a_2$, along with the cosine of the tilt angle of each component, 
$\cos{\theta_1}$ and $\cos{\theta_2}$. The tilt angle is defined as the angle between each components' spin vector and the binaries orbital angular momentum vector. 
We assume the polar spin angles are uniformly distributed in the orbital plane. While we choose to use M-Spline distributions to model the primary mass and spin distributions, we 
model the mass ratio distribution with a power law, i.e. $p(q | m_1, m_\mathrm{min}, \beta) \propto q^{\beta} \Theta(qm_1 - m_\mathrm{min}) \Theta(m_1 - qm_1)$, where $\beta$ is 
the power law index and $\Theta$ is the Heaviside step function which ensures $m_2$ is within the allowed range, [$m_\mathrm{min}$, $m_1$] \citep{Talbot_2018,o1o2_pop,o3a_pop}.
Additionally, we fit a population model on the redshift or luminosity distance distribution of BBHs, assuming a $\Lambda\mathrm{CDM}$ cosmology defined by the parameters 
from the Planck 2015 results \citep{Planck2015}. This defines an analytical mapping between each event's inferred luminosity distance, and it's redshift, which we now use interchangeably. 
We model the evolution of the merger rate with redshift with the \textsc{PowerlawRedshift} model, described by $p(z|\lambda_z)\propto \frac{dV_c}{dz}(1+z)^{\lambda_z-1}$ \citep{Fishbach_2018redshift}. 
We describe our specific prior choices on the hyperparameters and each model in more detail in Appendix \ref{sec:modelpriors}.

### 2022-10-09 src/tex/results.tex
Move some stuff around in prep for new results -- dont ever mention mspline

@@

We use hierarchical Bayesian inference (see Appendix \ref{sec:hierarchical_inference}) to simultaneously infer the astrophysical mass, spin and redshift distributions of 
binary black holes (BBHs) given a catalog of gravitational wave observations. We parameterize the binaries' masses with the primary (defined as the more massive component) mass 
and the mass ratio -- defined as $q=\frac{m_2}{m_1}$, bounded in the range from 0 to 1. Furthermore, we model 4 of the 6 total 
spin degrees of freedom of a binary merger with each component spin magnitude $a_1$ and $a_2$, along with the cosine of the tilt angle of each component, 
$\cos{\theta_1}$ and $\cos{\theta_2}$. The tilt angle is defined as the angle between each components' spin vector and the binaries orbital angular momentum vector. 
We assume the polar spin angles are uniformly distributed in the orbital plane. While we choose to use M-Spline distributions to model the primary mass and spin distributions, we 
model the mass ratio distribution with a power law, i.e. $p(q | m_1, m_\mathrm{min}, \beta) \propto q^{\beta} \Theta(qm_1 - m_\mathrm{min}) \Theta(m_1 - qm_1)$, where $\beta$ is 
the power law index and $\Theta$ is the Heaviside step function which ensures $m_2$ is within the allowed range, [$m_\mathrm{min}$, $m_1$] \citep{Talbot_2018,o1o2_pop,o3a_pop}.
Additionally, we fit a population model on the redshift or luminosity distance distribution of BBHs, assuming a $\Lambda\mathrm{CDM}$ cosmology defined by the parameters 
from the Planck 2015 results \citep{Planck2015}. This defines an analytical mapping between each event's inferred luminosity distance, and it's redshift, which we now use interchangeably. 
We model the evolution of the merger rate with redshift with the \textsc{PowerlawRedshift} model, described by $p(z|\lambda_z)\propto \frac{dV_c}{dz}(1+z)^{\lambda_z-1}$ \citep{Fishbach_2018redshift}. 
We describe our specific prior choices on the hyperparameters and each model in more detail in Appendix \ref{sec:modelpriors}. Following the same cut on the recent GWTC-3 catalog done in the LVK's 
accompanying BBH population analysis, we have 70 possible BBH mergers with false alarm rates less than 1 per year \citep{GWTC3,o3b_astro_dist}. Since it was concluded to be an outlier of the rest of 
the BBH population in both GWTC-2 and GWTC-3, we choose to omit the poorly understood event, GW190814 \citep{190814disc,o3a_pop,o3b_astro_dist,Essick_2022}. This leaves us with 
a catalog of 69 confident BBH mergers observed over a period of about 2 years, from which we want to infer its population properties. Our study provides the first comprehensive 
data-driven investigation of the BBH mass and spin distributions, uncovering new insights and corroborating those found with other methods. 
We start by looking towards our model's inferences of the mass distribution.

>>>>

We use hierarchical Bayesian inference (see Appendix \ref{sec:hierarchical_inference}) to simultaneously infer the astrophysical mass, spin and redshift distributions of 
binary black holes (BBHs) given a catalog of gravitational wave observations. Following the same cut on the recent GWTC-3 catalog done in the LVK's 
accompanying BBH population analysis, we have 70 possible BBH mergers with false alarm rates less than 1 per year \citep{GWTC3,o3b_astro_dist}. Since it was concluded to be an outlier of the rest of 
the BBH population in both GWTC-2 and GWTC-3, we choose to omit the poorly understood event, GW190814 \citep{190814disc,o3a_pop,o3b_astro_dist,Essick_2022}. This leaves us with 
a catalog of 69 confident BBH mergers observed over a period of about 2 years, from which we want to infer its population properties. Our study provides the first comprehensive 
data-driven investigation of the BBH mass and spin distributions, uncovering new insights and corroborating those found with other methods. 
We start by looking towards our model's inferences of the mass distribution.

@@
